---
title: "Replace Fuction"
author: "Alex Tidd"
format: html
---

```{r}
library(elo)
library(dplyr)
library(tidyverse)
library(cowplot)
library(ggrepel)
library(lubridate)
```

```{r}
NCAA_teams <- read_csv("NCAA_teams.csv")
schedule2324 <- read.csv("~/Desktop/SYE2024/schedule2324.csv")
```

```{r}
schedule2324 <- schedule2324 |>
  mutate(date = ymd(date))
```

```{r}
elo_ratings_initial <- tibble(home_team = schedule2324$home_team,
                              away_team = schedule2324$away_team,
                            elo_rating_home = NA,
                            elo_rating_away = NA,
                            outcome_home = schedule2324$outcome,
                            game_date = schedule2324$date) |>
  mutate(outcome_away = abs(outcome_home - 1))
```


```{r}
elo_main <- NCAA_teams |>
  mutate(rating = 1500) |>
  select(Team, rating)
```

```{r}
```{r}
schedule2324 <- schedule2324 |>
  mutate(date = ymd(date))
```

```{r}
elo_ratings_initial <- tibble(home_team = schedule2324$home_team,
                              away_team = schedule2324$away_team,
                            elo_rating_home = 1500,
                            elo_rating_away = 1500,
                            outcome_home = schedule2324$outcome,
                            game_date = schedule2324$date)
```

```{r}
k = 20
elo_ratings_test_alternative <- elo_ratings_initial |> 
  mutate(outcome_away = abs(outcome_home - 1)) |> 
  mutate(exp_home = if_else(game_date == "2023-10-07",
                            true = 1/(1 + 10^((elo_rating_away - elo_rating_home)/400)),
                            false = NA)) |>
  mutate(exp_away = if_else(game_date == "2023-10-07",
                            true = 1/(1 + 10^((elo_rating_home - elo_rating_away)/400)),
                            false = NA)) |>
  mutate(elo_new_home = if_else(game_date == "2023-10-07",
                                true = elo_rating_home + 20*(outcome_home - exp_home),
                                false = elo_rating_home)) |>
  mutate(elo_new_away = if_else(game_date == "2023-10-07",
                                true = elo_rating_away + 20*(outcome_away - exp_away),
                                false = elo_rating_away))
```

```{r}
elo_main$rating = replace(elo_main$rating, elo_main$Team, %in% elo_ratings_test_alternative$home_team)
elo_main$rating = replace(elo_main$rating, elo_main$Team, %in% elo_ratings_test_alternative$away_team)
```

